Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-11-17 MS/ms.Rmd
Update to MS to include link to supplement

We searched the Scopus database between May 16 and 18, 2021, with a follow up search on September 1, 2021 (see the Supplementary Information). After a series of pilot searches (detailed in Supplementary Materials) to refine our search string [See @Foo2021 for an overview of this process], we searched for studies that included: 'meta-analysis', 'meta regression', 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

@@

In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio, *$\ln{RR}$*. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., degrees Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question.

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random-effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals or from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various drivers of effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random-effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Weighted meta-analytic models are important to use because they: 1) improve the precision on meta-analytic mean estimates by accounting for sampling variance; 2) allow for a formal analysis of heterogeneity; 3) and give precedence to higher quality studies (see below). From our survey, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

As we have already emphasised, quantifying effect size heterogeneity provides the impetus for exploring why effects might vary within and across studies, while also providing context for interpreting the generality of overall effects. There are a number of measures commonly used to quantify effect size heterogeneity [@Borenstein2019]; however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

Heterogeneity described using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation; however, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Prediction intervals describe the range of plausible effect size values expected from a future study [@Nakagawa2021c]. This is different from a confidence interval that expresses the range of uncertainty around a statistical parameter estimate. Unlike $I^2$, prediction intervals (PI's) are probably more meaningful in the context of meta-analysis because they explicitly incorporate measures of dispersion around a mean effect size. They provide information about the likely effect size one can expect if we were to randomly sample a new population. For example, assume that we were interested in the overall impact of salinity stress in freshwater fish. To tackle this question, we collected experimental studies measuring fish swimming performance under high salinity treatments (\~15--20ppm Na Cl) compared to freshwater (\~0ppm Na Cl) controls, using *SMD* to quantify the effect salinity had on swim performance. We conduct a meta-analysis and estimate an overall *SMD* of 0.50 with a 95% prediction interval of 0.05 to 0.85. The PI indicates that effects can vary widely from as low as an *SMD* of 0.05 to as high as an *SMD* 0.85, depending on the population. In addition, if we were to repeat a similar study, we would expect a new effect to fall within the range of 0.05 and 0.85, 95% of the time [@Borenstein2019; @Nakagawa2021c]. This is a very intuitive interpretation of how variable effects are, yet PIs are often not reported in meta-analyses [@Nakagawa2021c], and we encourage more meta-analysts to report these (for examples of PIs, see Supplementary Materials).

@@

>>>>

We searched the Scopus database between May 16 and 18, 2021, with a follow up search on September 1, 2021 (see the Supplementary Information at https://daniel1noble.github.io/nuisance_hetreogeneity/). After a series of pilot searches (detailed in Supplementary Materials) to refine our search string [See @Foo2021 for an overview of this process], we searched for studies that included: 'meta-analysis', 'meta regression', 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Supplemental Tables). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

@@

In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio, *$\ln{RR}$*. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, if all the means are measured in the same units (e.g., degrees Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question.

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random-effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. Often, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals or from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various drivers of effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random-effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Weighted meta-analytic models are important to use because they: 1) improve the precision on meta-analytic mean estimates by accounting for sampling variance; 2) allow for a formal analysis of heterogeneity; 3) and give precedence to higher quality studies (see below). From our survey, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self-described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

As we have already emphasised, quantifying effect size heterogeneity provides the impetus for exploring why effects might vary within and across studies, while also providing context for interpreting the generality of overall effects. There are several measures commonly used to quantify effect size heterogeneity [@Borenstein2019]; however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

Heterogeneity described using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation; however, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Prediction intervals describe the range of plausible effect size values expected from a future study [@Nakagawa2021c]. This is different from a confidence interval that expresses the range of uncertainty around a statistical parameter estimate. Unlike $I^2$, prediction intervals (PI's) are probably more meaningful in the context of meta-analysis because they explicitly incorporate measures of dispersion around a mean effect size. They provide information about the likely effect size one can expect if we were to randomly sample a new population. For example, assume that we were interested in the overall impact of salinity stress in freshwater fish. To tackle this question, we collected experimental studies measuring fish swimming performance under high salinity treatments (\~15--20ppm Na Cl) compared to freshwater (\~0ppm Na Cl) controls, using *SMD* to quantify the effect salinity had on swim performance. We conduct a meta-analysis and estimate an overall *SMD* of 0.50 with a 95% prediction interval of 0.05 to 0.85. The PI indicates that effects can vary widely from as low as an *SMD* of 0.05 to as high as an *SMD* 0.85, depending on the population. In addition, if we were to repeat a similar study, we would expect a new effect to fall within the range of 0.05 and 0.85, 95% of the time [@Borenstein2019; @Nakagawa2021c]. This is a very intuitive interpretation of how variable effects are, yet PIs are often not reported in meta-analyses [@Nakagawa2021c], and we encourage more meta-analysts to report these (for examples of PIs, see Supplementary Material at https://daniel1noble.github.io/nuisance_hetreogeneity/).

@@

##Data Availability Statement
Supplementary information, data, code and results associated with this paper can all be found at: https://daniel1noble.github.io/nuisance_hetreogeneity/.
